---
title: "SpatCourse_diversity"
author: "Stefano Larsen"
date: "3/17/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, fig.width = 5.5, fig.height = 4, warning = F, message=F)
```

## load packages
```{r}
#install.packages('hillR')
library(hillR)

library(vegan)
library(tidyverse)
library(FD)

#install.packages('MeanRarity')
library(MeanRarity)

```

# Diversity measures

We explore how the *classic* diversity measures relate to the *modern* hill-numbers based metrics, which reflect the *effective number of species* in the community

We first simulate a community with known SAD (species abundance distribution), and calculate taxon richness, and classic Shannon, and Simpson diversity. Then we remove half of the taxa (randomly) and re-quantify diversity metrics, using both classic and hill-numbers based diversity metrics.

```{r}
# using the 'MeanRarity::fit_SAD' function we simulate community with 50 spp and known abundance distribution
community1=data.frame(t(fit_SAD(rich=50, simpson = 10)$rel_abundances)); colnames(community1)=paste("sp",1:50)

```

let's calculate classic Shannon and Simpson diversity
```{r}
diversity(community1, 'shannon')
diversity(community1, 'simpson')
```

Now, lets remove 50% of species, randomly [like a large disturbance]
Then calculate classic diversity for this halved community of 25 species 

```{r}
set.seed(1)
community2=sample(community1, 25); community2=sort(community2, decreasing = T)
diversity(community2, 'shannon')
diversity(community2, 'simpson')
```

While the richness has declined by 50%, Shannon and Simpson diversity showed only modest decline;

Let's try the same using the *effective number of species* hill-based diversity

Hill-Shannon
```{r}
renyi(community1, scale=1, hill=T) # this would be Shannon (q=1)
renyi(community2, scale=1, hill=T) 

```

So hill-Shannon declines by ~50% (from c.18 to c.9), mirroring the decline in species number;

Let' see Hill-Simpson

Hill-Simpson
```{r}
renyi(community1, scale=2, hill=T) # this would be Simpson (q=2)
renyi(community2, scale=2, hill=T) 
```
hill-Simpson also declined by ~50% (from c.10 to c.5).

### The Hill-based conversion linearises the relation between taxonomic richness and diversity, making them more comparable.

We can visualise this simulating a community with 100 species [with equal abundance = 1 ind]
```{r}
comm1=data.frame(t(rep(1,100))); colnames(comm1)=paste("sp",1:100)# simulated community of 100 spp

# calculate shannon classsic and hill-shannon for each species in the community
shannon.test=data.frame(matrix(ncol=5, nrow=100))# matrix to hold the values
shannon.test[,1]=seq(1,100,1)
for(i in 1:100) shannon.test[i,2]= diversity(comm1[1:i], 'shannon')
for(i in 1:100) shannon.test[i,3]= diversity(comm1[1:i], 'simpson')
for(i in 1:100) shannon.test[i,4]= renyi(comm1[1:i], scale=1, hill=T)
for(i in 1:100) shannon.test[i,5]= renyi(comm1[1:i], scale=2, hill=T)

names(shannon.test)=c ('richness', 'Shannon','Simpson', 'Hill_Shannon', 'Hill_Simpson')


```

Plot classic Shannon & Simpson vs Richness
```{r fig.width=5, fig.height=3}
par(mfrow=c(1,2))
plot(richness~Shannon, shannon.test, main='Shannon classic')
plot(richness~Simpson, shannon.test, main='Simpson classic')
```

Plot Hill-Shannon & Hill-Simpson vs Richness
```{r fig.width=5, fig.height=3}
par(mfrow=c(1,2))
plot(richness~Hill_Shannon, shannon.test, main='Hill Shannon')
plot(richness~Hill_Simpson, shannon.test, main='Hill Simpson')
```


## The same approach can be used for quantifying functional diversity
This can be done using the *'hillR'* package


First, verify that diversity from the 'hillR' package are the same as those calculated with vegan::renyi

```{r}
renyi(community1, scale=1, hill=T)
hill_taxa(community1, q=1)
```
Yes! GOOD; Now lets see how it works on dummy communities with functional traits as well.

### We can generate some fake trait values for the community1 and community2 species.
```{r}
set.seed(1)
comm1.trt=data.frame(trt1=rpois(100, 500), trt2=sample(1:100, 100), trt3=rgamma(100, 2,1))
row.names(comm1.trt)= colnames(comm1)# need to provide the same sp names on the trait data

#comm2.trt=comm1.trt[rownames(comm1.trt) %in% colnames(community2),] # subset traits for community2


```

Calculate the range of functional diversities provided by *hillR*, over each richness value
```{r}
# calculate some different functional diversity metrics; use q=1 to express it as hill-shannon
FD.test=data.frame(matrix(ncol=6, nrow=100))# matrix to hold the values
FD.test[,1]=seq(1,100,1)
for(i in 2:100) FD.test[i,2]= hill_func(comm1[1:i], comm1.trt[c(1:i),], q=1)[1]# Q
for(i in 2:100) FD.test[i,3]= hill_func(comm1[1:i], comm1.trt[c(1:i),], q=1)[2]# FDis
for(i in 2:100) FD.test[i,4]= hill_func(comm1[1:i], comm1.trt[c(1:i),], q=1)[3]# D_q
for(i in 2:100) FD.test[i,5]= hill_func(comm1[1:i], comm1.trt[c(1:i),], q=1)[4]# MD_q
for(i in 2:100) FD.test[i,6]= hill_func(comm1[1:i], comm1.trt[c(1:i),], q=1)[5]# FD_q
names(FD.test)=c ('richness', 'Q', 'FDis', 'D_q', 'MD_q','FD_q')

```

Let's plot the relation between number of taxa and functional diversities

```{r}
par(mfrow=c(2,2), mar = c(4, 4, 2, 2))

plot(richness~Q, data=FD.test)
plot(richness~FDis, data=FD.test)
plot(richness~FD_q, FD.test)
plot(richness~MD_q, FD.test)

```




Here the results of course depends on the trait values of the species .
However, different patterns emerge looking at classic functional diversity measure (Q, FDis) relative to hill-based diversity measures (D_q: functional hill number, the effective number of equally abundant and functionally equally distinct species;  FD_q: total functional diversity, the effective total functional distance between species of the assemblage). 
Refer to *Chiu & Chao (2014) PLoS ONE* for details on these metrics.


