---
title: "SpatCourse_SSN2"
author: "Stefano Larsen"
date: "4/8/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, warning = F, message=F)
```

Load libraries
```{r}
library(SSN)
```


# Tutorial with the ready-available spatial data in SSN R package
Note: part of these examples come from the original SSN vignette!

### In this script we will use the Middle Fork river data to fit models for mean summer water T. Then we will predict mean temperature across a range of un-sampled locations across the river network.

## Import the ssn file from the SSN R package system file
```{r}
mf04p <- importSSN(system.file("lsndata/MiddleFork04.ssn",
   package = "SSN"), predpts = "pred1km", o.write = TRUE)
```

### You can also import specific set of locations for which we want to make predictions.
This is done using the *importPredpts* comand.
```{r}
mf04p <- importPredpts(mf04p, "Knapp", "ssn")
mf04p <- importPredpts(mf04p, "CapeHorn", "ssn")

```

### We explore the mf04p SSN object.
It shows four groups of variables: the observed, and three sets of prediction locations.
```{r}
mf04p
```



### Creating distance matrices is necessary
```{r}
createDistMat(mf04p, predpts = "Knapp", o.write = TRUE,
	amongpreds = TRUE)
createDistMat(mf04p, predpts = "CapeHorn", o.write = TRUE,
	amongpreds = TRUE)
```


## Let's plot our data
```{r}
plot(mf04p, lwdLineCol = "afvArea", lwdLineEx = 6, lineCol = "blue",
   pch = 19, xlab = "x-coordinate (m)", ylab = "y-coordinate (m)",
   asp = 1)

```

## This chunk must be run all at once inside the .Rmd script.
Alternatively, run each line directly in the console.

### Here we first plot the rivers as spatial lines object. Then we plot the locations with observed data. Then we plot three sets of prediction points. The first is a diffuse 1-km apart prediction across the networks. The second (Knapp) and third (CapeHorn) are dense sets of prediction points used for *block-kriging'.
```{r}
#plot as spatial lines object
plot(as.SpatialLines(mf04p), col = "blue",
	lwd = 1+ log(1+ mf04p@data$afvArea)*6)

# add the observed locations with size proportional
# to mean summer temperature
plot(as.SpatialPoints(mf04p), pch = 19,
cex = as.SpatialPointsDataFrame(mf04p)$Summer_mn/15 , add = TRUE)

# add the prediction locations on the 1 km spacing
plot(as.SpatialPoints(mf04p, data = "pred1km"),
	cex = 1, add = TRUE, pch = 1)

# add the dense set of points for block prediction on Knapp segment
plot(as.SpatialPoints(mf04p, data = "Knapp"), pch = 19, cex = 0.3,
  col = "red", add = TRUE)
# add the dense set of points for block prediction on CapeHorn segment
plot(as.SpatialPoints(mf04p, data = "CapeHorn"), pch = 19, cex = 0.3,
  col = "Gold3", add = TRUE)
```

## We can have a look at the values of mean summer T. 
Use *nclasses* to create break points for plotting.
```{r}
plot(mf04p, 'Summer_mn',, lwdLineCol = "afvArea", lwdLineEx = 6, lineCol = "blue",
   pch = 1, xlab = "x-coordinate (m)", ylab = "y-coordinate (m)",
   asp = 1, nclasses=8, main='Summer T')
```

### As always is good to have a look at the Torgegram.
We see that autocorrelation in water Tis high among flow-connected locations (lower variances), while flow-unconnected locations have larger variances.
```{r}
tor.summer_mn<- Torgegram(mf04p, "Summer_mn", nlag = 20, maxlag = 40000)
plot(tor.summer_mn)
```


# We can now fit some models

### We start by fitting a non-spatial models for summer T using elevation and slope as covariates
The summary shows that both elevation and slope are important covariates, explaining summer T.
```{r}
mf04.glmssn0 <- glmssn(Summer_mn ~ ELEV_DEM + SLOPE, mf04p,
   CorModels = NULL, use.nugget = TRUE, EstMeth = "ML")
summary(mf04.glmssn0)
```


### Then we can fit a spatial models including all autocovariance functions.
The summary shows that, in the spatial model, slope is not significant anymore. This happens often that spatial models have less frequent significant predictors.
```{r}
mf04.glmssn1 <- glmssn(Summer_mn ~ ELEV_DEM + SLOPE, mf04p,
   CorModels = c("Exponential.tailup", "Exponential.taildown",
      "Exponential.Euclid"), addfunccol = "afvArea", EstMeth = "ML")
summary(mf04.glmssn1)

```
### We can fit another model excluding the effect of slope.
```{r}
mf04.glmssn2 <- glmssn(Summer_mn ~ ELEV_DEM , mf04p,
   CorModels = c("Exponential.tailup", "Exponential.taildown",
      "Exponential.Euclid"), addfunccol = "afvArea", EstMeth = "ML")

summary(mf04.glmssn2)

```

```{r}
varcomp(mf04.glmssn2)
```


### We fit a third model, this timee exlcuding the Euclidean autocovariance function.
```{r}
mf04.glmssn3 <- glmssn(Summer_mn ~ ELEV_DEM , mf04p,
   CorModels = c("Exponential.tailup", "Exponential.taildown"), addfunccol = "afvArea", EstMeth = "ML")
summary(mf04.glmssn3)
```


## We compare the different models in terms of RMSPE, AIC and COV.90
It looks like *mf04.glmssn1* is the best model among these.
```{r}
InfoCritCompare(list(mf04.glmssn0,mf04.glmssn1,mf04.glmssn2, mf04.glmssn3 ))[,c(3,5,8,12)]
```

## Let's explore the residuals
The result of the residuals function is an `influenceSSN' object, which is
an exact copy of the `glmssn' object, except that residual diagnostics are appended as new
columns to the data frame point.data containing the observed data. The default plotting
method for an `influenceSSN' object is a map with color-coded raw residuals.
```{r}
# get the residuals from the model
mf04.resid1 <- residuals(mf04.glmssn1)
# Explore the new variables appended to the data
names(getSSNdata.frame(mf04.resid1))
#simple plot of residuals
plot(mf04.resid1)
```

### The residual plot shows that some locations have rather large residual values (< -3). We could eventually remove these outliers. For now we carry on using all the data points.


# We can now make predictions of summer T, using the model we have fitted.
## First we predict over the range of locations across the networks. These are the points data *pred1km*.
```{r fig.height=4}
mf04.pred1km <- predict(mf04.glmssn1, "pred1km")
plot(mf04.pred1km, SEcex.max = 1, SEcex.min = .3, nclasses=5)

```

## Then we can also make block predictions, over a range of adjacent points along a river reach. here are the points along the *Knapp* river.

```{r}
# Plot zooming to the 'Knapp' river where predictions are needed
plot(mf04p, "Summer_mn", pch = 1, cex = 3,
   xlab = "x-coordinate", ylab = "y-coordinate",
   xlim = c(-1511000,-1500000), ylim = c(2525000,2535000))

#Run the predictions using the 'Knapp' prediction locations
mf04.glmssn1.Knapp <- predict(mf04.glmssn1, "Knapp")

#The plot it
plot(mf04.glmssn1.Knapp, "Summer_mn", add = TRUE,
   xlim = c(-1511000,-1500000), ylim = c(2525000,2535000))
```

$$\\[1.1in]$$

# Excercise


## We have modelled water temperature using gaussian distribution (the default).

## However, the glmssn allows different distribution of the response variable, including binomial and poisson (counts).

The mf04p dataset also contains a variable (*C16*) reporting the number of days the water T was above 16C. This is therefore a *count* variable that can be modelled using a poisson distribution. 

### Here is the variable *C16*
```{r}
mf04p@obspoints@SSNPoints[[1]]@point.data$C16
```

## You can try yourself:

* Fit a poisson model for the variable *C16* using again elevation and slope as covariates.
* Exclude non-significant predictors.
* Predict the *C16* over the prediction locations (e.g. Knapp stream) using your model.


## TIP
In SSN we can include the line: family = "poisson" in the *glmssn* call,  as in standard *glm* models.

Then you can use the examples already provided to make prediction on the Knapp stream.


