---
title: "SpatCourse_SSN1"
author: "Stefano Larsen"
date: "3/22/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, fig.width = 5.5, warning = F, message=F)
```


# Scripts and examples for visualising and analysing Spatial Stream Network models (SSN)

These scripts provide introductory examples on how to deal with .ssn objects in R.
We will work with a ready-available .ssn object in .RData format. You can download this from:
*https://github.com/stefanolarsen/SpatCourse*
The dataset include macroinvertebrates community data across the Adige River network (NE Italy). Environmental descriptors, diversity and feeding-trait metrics are included. For details see:
**Larsen et al 2019 - Testing the river continuum concept with geo-statistical stream network models**. *Ecological Complexity.* https://doi.org/10.1016/j.ecocom.2019.100773

## Load the key libraries.
```{r}
library(SSN)
library(tidyverse)
#library(gtools)
```


## Load the RData with the SSN object (stream invertebrates from Adige River network)
You must first download this .RData object from the *https://github.com/stefanolarsen/SpatCourse*.
Then modify the code below to include your path (where you downloaded the file).
```{r}
load("~/Documents/SpatCoursePT/SpatCourse/spat.course.ssn1.RData")

#gitUrl='https://github.com/stefanolarsen/SpatCourse/blob/main/bent.ssn.logit.RData'
```



## Create the distance matrices. 
These are non-orthodox distance matrices. See help for details.
Distance matrix is important for modelling later!
```{r}
createDistMat(spat.course.ssn)
```

## Now that you have the .ssn object imported, you can explore it.

Spatial objects can be explored with *summary*.

```{r}
summary(spat.course.ssn)
```


## Explore the variables in the data only.
Here there are diversity metrics (richness, shannon).
Other descriptors e.g. T_wat (water T), LIMeco (water quality index).
Other variables (netID, ratio) come from the process leading to the SSN file, from GIS.
## An important variable is *afvArea* (the additive function based on catchment area), use for weighting the correlation function at river juntions.
```{r}
names(spat.course.ssn)

```

## Explore plotting functions

Simple *plot* function can also work. You see the network and the sampling points.
```{r}
plot(spat.course.ssn, asp=1, cex=0.5)
```


### Nicer option for plotting; can weight the line width by catchment area.
The option here is: 'lwdLineCol=afvArea'.
```{r}
plot(spat.course.ssn, lwdLineCol = "afvArea", lwdLineEx = 6, lineCol = "blue",
   pch = 1, xlab = "x-coordinate (m)", ylab = "y-coordinate (m)",
   asp = 1)

```

## Exploring patterns in specific variables

You can also plot specific variable values, with color classes (defined in 'nclasses'). 
Here e.g. the % of agricultural land-use around each site (buf.agr.p).
```{r}
plot(spat.course.ssn, 'buf.agr.p',, lwdLineCol = "afvArea", lwdLineEx = 6, lineCol = "blue",
     pch = 1, xlab = "x-coordinate (m)", ylab = "y-coordinate (m)",
   asp = 1, nclasses=5, main='% agric landuse')

```

Or plot the water temperature (T_wat).
```{r}
plot(spat.course.ssn, 'T_wat',, lwdLineCol = "afvArea", lwdLineEx = 6, lineCol = "blue",
   pch = 1, xlab = "x-coordinate (m)", ylab = "y-coordinate (m)",
   asp = 1, nclasses=8, main='water T')

```



# Manipulate the data frames inside a SSN object
In this example you need to:
e.g. Fill NAs in the water quality index: LIMeco
Re-transform "logit"-transformed traits to proportions for some traits values.


## Extract dataframe from ssn object.
This command extract the data matrix from the ssn object. 
We can work on it, and then put it back into a ssn object.  This is useful if we need to modify or add data to our dataframe.
```{r}
datassn=getSSNdata.frame(spat.course.ssn)
```

## Work on the dataframe. Explore structure.
This is now like a standard dataframe in R.

```{r}
str(datassn)
#some feeding traits proportions are expressed as logit (e.g feed_xyl)
#we can convert it back to proportions if we want.
```

Look for missing values
```{r}
sum(is.na(datassn$LIMeco)) 
# there are missing values in LIMeco
```

We can fill missing values in LIMeco with overall mean.
```{r}
datassn$LIMeco[is.na(datassn$LIMeco)]<- mean(datassn$LIMeco, na.rm=T)# 
```



Just a function to transform logit back to proportion of some feeding traits.

*delogit function*: convert logit transformed proportions back to proportions
```{r}
delogit=function(x){
  x1=exp(x)
  prop=x1/(1+x1)
  return(prop)
}
```

Convert these traits back to proportions.
```{r}
datassn$feed_oth=delogit(datassn$feed_oth)
datassn$feed_xyl=delogit(datassn$feed_xyl)
```



## Then we create a second SSN object with the new modified dataframe.
We can call it the same way for convenience (overwrite the existing one).
```{r}
spat.course.ssn=putSSNdata.frame(datassn, spat.course.ssn)
#save(spat.course.ssn, file='SpatCourse.ssn.RData')
```

To specifically explore the data within a SSN object:access the slots using *@*
```{r}
head(spat.course.ssn@obspoints@SSNPoints[[1]]@point.data)
```
$$\\[0.9in]$$


## Exploring variograms for river networks. the *Torgegram*


Explore default *torgegrams*.
SSN package offers this default plots for exploring variograms.
It shows the variance for flow-connected and flow-unconnected observation pairs.
The size of the spots reflect the #observation pairs for each distance lag.
Here an example of torgegram for 'shannon diversity'. *Clearly different spatial patterns for flow-connected and unconnected locations.*
```{r}
tor.shn=Torgegram(spat.course.ssn, 'shannon', nlag = 50, maxlag = 400000)
plot(tor.shn)
```


### Let's see how the torgegram object looks like.
It shows the distance lags, the variance ('gam') and the n. of observation pairs within the lag ('np').
And it does so for flow-connected and unconnected locations.
```{r}
tor.shn
```



# Variograms can be better examined including the three distance types.
Exploring the shapes of variograms over the Euclidean, flow-unconnected and flow-connected dimensions provide more insight into the main spatial patterns of the data.


## Let's examine the variogram for water T, along the three distances.
```{r}
# first the classic variogram over Euclidean dist
emp.var.wt=EmpiricalSemivariogram(spat.course.ssn, 'T_wat', nlag = 30, maxlag = 200000)
#then the torgegram for flow-connected and unconnected dist
tor.wt=Torgegram(spat.course.ssn, 'T_wat', nlag = 50, maxlag = 400000)
```

Plot all distances together
```{r}
plot(tor.wt$gam.unconnect[1:20]~tor.wt$distance.unconnect[1:20], options(scipen = 5), 
     type='b', lty=1, pch=19, ylim=c(0,15), col='blue', ylab='Semivariance T_wat', 
     xlab='Distance (m)', cex.lab=1, las=1)

lines(tor.wt$gam.connect[1:13]~tor.wt$distance.connect[1:13], type='b', col='purple', 
      lty=3, pch=19)
lines(emp.var.wt$gamma[1:23]~emp.var.wt$distance[1:23],col='orange',  type='b', 
      pch=19, lty=3 )
legend(100, 14, legend=c("Eucl", "F-con", 'F-unc'),
       col=c("orange", "purple", 'blue'), lty=c(1,1,1), lwd=2,  cex=0.7)


```

### The above plot shows that, for a given distance, water T varies little over the watercourse dimension, and more over the Euclidean dimension, but only beyond 50 km distance.

## Now let's examine the variogram for local land-use
```{r}
# first variogram along the Euclidean dimension
emp.var.agr=EmpiricalSemivariogram(spat.course.ssn, 'buf.agr.p', nlag = 30, maxlag = 200000)
# then the torgegram over flow-connected and unconnected distances
tor.agr=Torgegram(spat.course.ssn, 'buf.agr.p', nlag = 50, maxlag = 400000)
```

Plot it together
```{r}
plot(tor.agr$gam.unconnect[1:20]~tor.agr$distance.unconnect[1:20], options(scipen = 5), 
     type='b', lty=1, pch=19, ylim=c(0,1600 ), col='blue', ylab='Semivariance agr. landuse', 
     xlab='Distance (m)', cex.lab=1, las=1)

lines(tor.agr$gam.connect[1:13]~tor.agr$distance.connect[1:13], type='b', 
      col='purple', lty=3, pch=19)
lines(emp.var.agr$gamma[1:23]~emp.var.agr$distance[1:23],col='orange',  
      type='b', pch=19, lty=3 )
legend(100, 1500, legend=c("Eucl", "F-con", 'F-unc'),
       col=c("orange", "purple", 'blue'), lty=c(1,1,1), lwd=2,  cex=0.7)

# Here we see that local landuse varies more rapidly along flow-connected locations
# reflecting the changes from headwaters to mainstems
```

### The above variogram show that agricultural landuse changes rapidly along flow-connected locations, reflecting changes from up to downstream within ~50km.


## As extreme case, we can explore the variogram for the variable *PCA1.long*. This is a synthetic variable refelecting the logitudinal continuum. As expected, the variance increases along flow-connected locations, but remains low over the other dimensions.
```{r}
# first the variogram for Euclidean and then the Torgegram
emp.var.PCA1=EmpiricalSemivariogram(spat.course.ssn, 'PCA1.long', nlag = 30, maxlag = 200000)
tor.PCA1=Torgegram(spat.course.ssn, 'PCA1.long', nlag = 50, maxlag = 400000)
```
Plot it together as before
```{r}
plot(tor.PCA1$gam.unconnect[1:20]~tor.PCA1$distance.unconnect[1:20], options(scipen = 5), 
     type='b', lty=1, pch=19, ylim=c(0,8), col='blue', ylab='Semivariance PCA1', 
     xlab='Distance (m)', cex.lab=1, las=1)

lines(tor.PCA1$gam.connect[1:13]~tor.PCA1$distance.connect[1:13], type='b', col='purple', 
      lty=3, pch=19)
lines(emp.var.PCA1$gamma[1:23]~emp.var.PCA1$distance[1:23],col='orange',  type='b', 
      pch=19, lty=3 )

legend(100, 7, legend=c("Eucl", "F-con", 'F-unc'),
       col=c("orange", "purple", 'blue'), lty=c(1,1,1), lwd=2,  cex=0.7)
```

$$\\[0.7in]$$


# Now let's run some models. Explore the influence of key environmental drivers on macroinvertebrate metrics.
## First, have a look again at the variables in the data; 
```{r}
names(spat.course.ssn)
```


### As example, we fit models for the porportion of shredders (*feed_shr*) in the community. 
### We will include the longitudinal gradient (*PCA1.long*), the % of agricultural landuse (*buf.agr.p*) and water quality (*LIMeco*), as predictor (covariates).

#### Start with a non-spatial model (CorModels=NULL). This is equivalent to a classic lm.
```{r}
shr.0=glmssn(feed_shr~PCA1.long+buf.agr.p+LIMeco,spat.course.ssn,
             CorModels = NULL, use.nugget = TRUE,EstMeth = "ML")             

summary(shr.0)
```

We see that both landuse and water quality are important factors.



### Now we can fit some spatial models. We start with the full autocovariance functions (Euclidean, Tail.up, Tail.down).
```{r}
shr.1=glmssn(feed_shr~PCA1.long+buf.agr.p+LIMeco,spat.course.ssn,
             CorModels = c("Exponential.tailup", "Exponential.taildown",
   "Exponential.Euclid"), addfunccol = "afvArea", EstMeth = "ML")

summary(shr.1)
```

Here we see a different output. The variogram parameters for each autocovariance functions are also given. We also see that the model parameters (*Estimate, Std.Error*) are different from the non-spatial model. We also see that the Sill of the tail.down component is very small. We could eventually omit it from the model.


## Model selection is a stepwise process. Once we are happy with the covariates (the fixed effect), we can work on the autocorrelation structures. 
### Here we fit a couple of additional models with different variogram functions.


```{r}
shr.2=glmssn(feed_shr~PCA1.long+buf.agr.p+LIMeco,spat.course.ssn,
             CorModels = c( "LinearSill.tailup",
   "Exponential.Euclid"), addfunccol = "afvArea", EstMeth = "ML")

#summary(shr.2)
```

```{r}
shr.3=glmssn(feed_shr~PCA1.long+buf.agr.p+LIMeco,spat.course.ssn,
             CorModels = c( "Exponential.tailup"), addfunccol = "afvArea", EstMeth = "ML")

#summary(shr.3)
```


```{r}
shr.4=glmssn(feed_shr~PCA1.long+buf.agr.p+LIMeco,spat.course.ssn,
             CorModels = c( "Exponential.Euclid"), addfunccol = "afvArea", EstMeth = "ML")

#summary(shr.4)
```


### We can then use the command *InfoCritCompare* to compare the different models for e.g. Root Mean Square Prediction Error (RMSPE) and AIC.
```{r}
InfoCritCompare(list(shr.0, shr.1, shr.2, shr.3, shr.4))
```

### We can print specific values for clarity (e.g. Spatial components, AIC, RMSPE)
```{r}
InfoCritCompare(list(shr.0, shr.1, shr.2, shr.3, shr.4))[,c(3,5,8,12)]
```

## We see that spatial models including tail-up and/or tail-down (accounting for river topology) provide better predictions (lower RMSPE) than simple non-spatial model (shr.0), or models that only include Euclidean component (shr.4).


### The *varcomp* function provide information on the variance components of the model (covariates + spatial components)
```{r}
varcomp(shr.2)
```

In the model *shr.2*, the included covariates (PCA1.long, Limeco, Buf.agr.p) explained 34% of variation. The tail-up autocovariance (reflecting correlation along flow-connected locations) accounted for 13% of variation, while the Euclidean covariance for 27%. The Nugget reflects the unexplained variance (with no spatial structure).


### We can also plot the residuals of the model easily; this helps visualising spatial patterns in the residuals.
```{r}
plot(resid(shr.2), asp=1, cex=0.5)
```




$$\\[0.7in]$$


# Excerice that you can try yourself:

* Plot additional variograms for other variables (e.g. Limeco, Shannon)
* Fit models for the proportion of predators (*feed_pre*) using the same covariates used before.
* Try to fit both non-spatial and spatial models (with different covariance functions).
* Compare the models using *InfoCritCompare*.


